Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-10-16 manuscript/manuscript.tex
New images. Review of about one-half of manuscript
Minor code changes. Updated charts in the image directory.  Editing Manuscript.

\usepackage{comment}
\usepackage{booktabs, textgreek}

@@

The paper is organized as follows. Section~\ref{sec:data} provides 
an overview of the history of the NYC 311 SR system and presents 
a summary of SR counts over the 10-year period from 2014 to 2023. 
Section~\ref{sec:issues} delves into specific data cleansing 
challenges affecting data quality and curation efficiency, 
including structural problems, adherence to the data dictionary, 
and the presence of missing, blank, or N/A entries. This section
utilizes 311 SR data over a a 2-year period from 2022-2023.  It 
 investigates field compliance with reference or acceptable values, 
highlights logical inconsistencies, and examines concerning patterns 
in the data. We explore the balance between precision and accuracy 
and identify duplicate or redundant fields, along with observations 
on the Data Dictionary. Section~\ref{sec:recommendations} offers 
practical recommendations for mitigating or resolving these issues, 
while Section~\ref{sec:discussion} encapsulates key insights and 
discusses the broader implications of our findings.

\section{NYC 311 SR Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023.

@@

\begin{comment}
Despite its success, the 311 system faces several challenges:
data timeliness, accuracy, and consistency, difficulties in
correlating data over long periods, excluding personally
identifiable information (PII), integration with stand-alone systems
at selected NYC agencies, and managing API usage for numerous
third-party users. These challenges are addressed by various agency
open data managers and 
\end{comment}

@@

\begin{figure}[tbp]
	\centering
  	\includegraphics[width = \textwidth]{10-year-trend_yearly.pdf}
 	\caption{10-year (2014-2023) annual SR Counts.}
  	\label{fig:10-yr}
\end{figure}

Our investigation utilizes two 311 SR datasets covering 10-years and 2-years.
The 10-year period (2014--2023), downloaded on 15 September 2024, 
contains only three columns (created\_date, closed\_date, and 
agency). This dataset is used to analyze long-term 
trends and observe certain dating anomalies. The 10-year dataset is approximately 
1.4 GB in size and contains over 27 million rows, each row representing one SR.
 Instructions on downloading this file from the NYC Open Data Portal 
 are available in the supplemental material.

Additionally, this investigation uses a 2-year dataset covering the period
2022--2023, also downloaded on 15 September 2024. This data set 
contains the full 41 columns of 311 SR data available from the portal, 
and is used to conduct detailed analysis in areas such as data consistency, 
data validity, redundant fields, etc.. This 2-year dataset 
is quite large at 3.7GB. It contains over 6 million rows of SRs. Instructions 
on downloading this file from the NYC Open Data Portal are 
available in the supplemental material.

Figure~\ref{fig:10-yr} shows  annual counts of 311 SRs over the
10-year period from 2014 to 2023, revealing a clear upward trend in
public engagement. As indicated, the most significant increase
in activity occurred in 2020, when the number of requests surged to 3.23 million
during the COVID-19 pandemic, illustrating the essential role of the
311 system in supporting residents during times of crisis. Overall, SR
counts increased by 49.5\% between 2014 and 2023, reflecting
growing public reliance on the system. (During this 10-year time 
period, the population of NYC grew just 0.6\%.) The steady 
rise in SRs can be attributed to the increased accessibility of 
the 311 system via online and mobile platforms, as well as 
heightened public awareness  and utilization of the service. 
While the spike in 2020 was exceptional, the broader trend 
indicates sustained growth in service request volume over 
the decade, highlighting the system's evolving role in managing 
both routine city operations and extraordinary events.

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
agency, showing the cumulative percentage of SRs handled by each
agency over a the 2-year time frame. Note that the
distribution of SRs is concentrated among a few key agencies. The six
largest agencies are New York Police Department (NYPD), 
Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).
Collectively they handle over 90\% of the total SR volume. This is
indicative of the critical roles these agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The NYPD alone accounts for a
substantial portion of the total SRs, underscoring its central role in
addressing complaints related to public safety. The
remaining 10\% is distributed across a number of other
agencies. This concentration of SRs points to a
reliance on a few key departments to resolve most non-emergency
issues (e.g. non 911 calls). Accordingly, this underscores 
the necessity for optimal data handling and curation processes 
within these high-volume agencies in order to ensure 
efficiency and responsiveness.

@@

Figure~\ref{fig:SR_complaints}
illustrates the top 20 complaint types, accounting for 70\% 
of all SRs and their cumulative percentage. Note that the 
distribution of complaints is skewed, with a small number of issues 
dominating total SR volume. Noise-related complaints, including 
residential, commercial, and street noise, make up 22\% of all 
complaints, making them the most frequent issue in the NYC 311 system. 
Other prominent categories include illegal parking, heat/hot water 
complaints, and blocked driveways. The cumulative percentage curve 
shows that after the top 20, the remaining complaint types are spread 
thinly across the remaining 30\% of SRs. This concentration suggests 
that improving responses to the most frequent complaints could have 
a significant impact on overall service efficiency and resident 
satisfaction. It also provides insights into the operational 
pressures faced by the city agencies responsible for handling these 
high-volume complaints.

@@

Data cleansing refers to the process of identifying and rectifying
errors, inconsistencies, and inaccuracies within datasets to ensure
they are of high quality and reliable for analysis
\citep{maletic2005data, hosseinzadeh2023data}. The process
typically involves removing duplicate records, handling missing or
incomplete data, correcting mislabeled or inaccurate entries, and
standardizing data formats \citep[e.g.,][]{cody2017cody,
  van2018statistical}. In the context of open data, cleansing is
especially important as open datasets often come from diverse,
uncoordinated sources, leading to variations in data quality,
completeness, and consistency. Without thorough data cleansing, the utility
of open data can be quite limited, affecting its reliability for
research, policy-making, and innovation. The main purpose of cleansing
open data is to ensure that it is accurate, consistent, and usable
across multiple platforms and by various stakeholders. This improves
the trustworthiness of the data and enables better decision-making,
more accurate analysis, and the integration of data into machine
learning models or other systems.

Many quality criteria are employed to ensure high-quality data 
processing. One of the primary efforts is data validation, which spans 
several critical checks. For instance, mandatory fields must not be 
left empty, ensuring that key information is always captured. 
Additionally, certain fields must conform to specific data types, 
such as numeric, character, or date formats, which are typically 
outlined in a Data Dictionary. Another essential aspect of validation 
is domain compliance, where fields must adhere to a predefined set of 
values, such as statuses, state names, zip codes, or gender. Structural 
errors also play a significant role in data quality, particularly when 
naming conventions or data entries are inconsistent. Common issues 
include fields that do not appear in the Data Dictionary or the 
inconsistent use of blanks, spaces, NA, N/A, or ``<NA>'' to indicate 
missing data. Furthermore, redundant or irrelevant fields can clutter 
datasets, reducing efficiency. Logical inconsistencies are another 
important consideration, such as related fields that violate expected 
relationships, such as a ``due date'' that precedes the ``created date.'' 
Lastly, the balance between accuracy and precision is crucial, as both 
must be carefully managed to ensure reliable and meaningful data.

Here we identify the presence of issues in the NYC 311 SR open data;
not attempting to solve them. A solution effort should
be undertaken only after an investigation as to the why and how 
such issues came about, and a discussion as to whether or not it 
is an actual \textit{error} or holds some other status.

@@

	\item There are four date fields (created, closed, updated, due).

@@

	\item In addition to the incident\_address, there are five additional location fields: 
	lat/long, street\_name, landmark, NY State plane, and Block \#

	\item One free-form text field, resolution\_description, which 
	supports 934 characters of input, including commas and special characters
\end{itemize}

The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} 
identifies 41 data columns (fields) and provides related information 
for each. However, there are six additional fields that can be
optionally downloaded. These fields are 
visible in the portal's ``Column manager'' widget, labeled with the prefix 
``@computed\_region.'' The additional fields include \texttt{zip\_codes}, 
\texttt{community\_districts}, \texttt{borough\_boundaries}, 
\texttt{city\_council\_districts}, \texttt{police\_precincts}, and 
\texttt{police\_precinct}. While one can likely infer the meaning and 
derivation of these computed fields, uncertainty remains as to
their derivation and usage. These six fields are not included in
the Data Dictionary. Our analysis has shown that these six 
fields have numerous data validity issues. Liaison with the 
NYC Open Data Teams leads us to treat these six fields 
as experimental and not for suitable for normal use. Accordingly, 
we will not address these fields in our analysis.

\subsection{Missing Data}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to 
determine if SRs were closed before or after their
\texttt{due\_date}, you would be challenged as 99.6\% of the
\texttt{due\_date} field is
blank. When counting fields for blank or N/A values, they appear 
to divide into three groups:
\textbf{Mostly Empty}, \textbf{Partially Empty}, 
and \textbf{Few/None Empty}. 
The Mostly Empty category ranges from 93--99.9\% blank 
and includes such fields as
\texttt{taxi\_company due\_date},
\texttt{pickup\_location}, and \texttt{landmark}.
The Partially Empty includes such fields as
\texttt{location\_type}, \texttt{borough}, 
and \texttt{cross\_street}. And the Few/None Empty includes
\texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. In some cases it may make
sense to inquiry as to why some fields are frequently blank.
Figure~\ref{fig:blank_fields} presents a graphic 
depiction of total empty (blank \& N/As) for each fields illustrating
the grouping into the Most, Partial, and Few categories.

@@

\subsection{Validating Data for Acceptable Values}
\label{sec:domain}
Any analytic effort must ensure that fields containing invalid values 
are identified and removed from the analysis. Below are the results of 
selected field validation. The latitude and longitude fields were 
found to fall within the geographic boundaries of New York City, and 
the \texttt{unique\_key} field was indeed unique, as required. 
Unfortunately, the Data Dictionary specifies very few domains of 
acceptable values. However, the following fields were tested and found 
to comply with their expected domains, as determined by public usage 
and historical datasets: \texttt{address\_type}, \texttt{status}, 
\texttt{borough}, \texttt{borough\_boundaries}, \texttt{park\_borough}, 
\texttt{data\_channel}, \texttt{vehicle\_type}, and 
\texttt{city\_council\_district}. Some fields proved to be problematic 
as to their compliance with a domain of legal values.

\paragraph{Zip Codes}
\label{sec:zipcodesissues}
All zip codes ( \texttt{incident\_zip}) 
should be valid as defined by the USPS database, which contains 
44,173 valid zip codes. The \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,163 errors; 
not insignificant. If you sort the invalid entries by Agency, 
the distribution by percentage mirrors the \texttt{overall} 
breakdown of SRs by Agency, potentially 
indicating a systemic problem.

\paragraph{Date and Closed Date}
\label{sec:negativeduration}
For the \texttt{created\_date} and \texttt{closed\_date} fields, one 
might expect that these fields are automatically populated by the SR 
application software when setting an SR status to ``new'' or ``closed.'' 
Unfortunately, this does not seem to be the case, as several anomalies 
exist in the date fields:
\begin{itemize}
    \item SRs with a \texttt{closed\_date} that occurs before the 
    \texttt{created\_date}.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} in 
    the far distant past.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} that 
    match to the second.
    \item A large number of SRs closed and/or created exactly at midnight 
    or noon, to the second.
\end{itemize}

\subparagraph{Closed before Created: Negative Duration}
Citizens, NYC Government Officials, and Agencies use the created and 
closed dates to measure the ``duration'' of SRs, which reflects Agency 
responsiveness. While duration is not directly present in the dataset, 
it can be easily computed as the difference between
\texttt{(closed\_date} and \texttt{created\_date}.
There are 12,251 SRs where the \texttt{closed\_date} precedes the 
\texttt{created\_date}, generating nonsensical ``negative durations.'' 
Though this represents only 0.2\% of the dataset, these errors can 
significantly impact response time analyses.
Eight SRs with extremely large negative durations ($-$730 days), all 
originating from the Department of Homeless Services (DHS), contain 
an entry of ``1900-01-01'' as the \texttt{closed\_date}. This results 
in negative durations exceeding $-$44,601 days (or 122 years). Such 
anomalies, though rare, can significantly skew statistical results if 
not addressed during data cleansing. As a result, these SR rows are 
removed from the analysis.

@@

@@

Excluding the extreme negative values, 
Figure~\ref{fig:negative-duration-violin} shows the broad spread of 
negative-duration SRs. While there are few outliers, the magnitude 
of the negative durations is troubling and can produce bizarre 
analytical results. Table~\ref{tab:largest-errors} summarizes the 
negative durations of the largest magnitude (in days). The data 
suggests that the negative duration issue is predominantly a problem 
within the Department of Transportation (DOT), where 95\% of these 
errors occur.

% For illustration, consider a case study of homeless person
% assistance. Suppose that DHS wants to know ``How 
% quickly have 311 SRs for ``Homeless Person Assistance'' been resolved 
% over the past two years?'' This is a typical request made by 
% both the public and City government; analogous to ``How quickly 
% is my Agency responding to ``critical'' requests, and does 
% that performance vary by Borough, Zip Code, etc. To conduct such an 
% analysis, we can use the tools on the Open Data Portal. Here are the steps:

% \begin{itemize}
%     \item Select created\_date from 2022-2023 and filter by complaint\_type = 
%     ``Homeless Person Assistance'' (this yields 55,000 SRs)

%     \item Compute the ``duration'' (closed\_date – created\_date)

%     \item Take an average of the ``duration'' field  which yields: \textbf{Answer:  -4.8 days}  
% \end{itemize}

% Clearly that answer is nonsensical. How did such a simple task result 
% in an absurd answer? The answer lies in the computation of the ``duration'' 
% field. It turns out, there are eight DHS SRs that have a closed\_date 
% of ``1900-01-01''. Each of those SRs creates a negative duration of -44,602 
% days (-122 years). Those eight SRs are enough to drive the 
% average duration of 55,000 Homeless Assistance SRs to a negative 
% value; clearly incorrect. In this case, the median is a better measure 
% of central tendency. The median is  0.2 days (approx 5 hrs) which will
% surely please the DHS.

%\begin{figure}[tbp]
% 	 \centering
 %	 \includegraphics[width = \textwidth]{homeless_response_time_clean.pdf}
%	 \caption{Homeless Assistance SR Durations}
%	 \label{fig:homeless}
%end{figure}

@@

@@

A similar issue occurs when resolution action update date is after the
closing date. \jy{It might be possible, I think, if they allow updates
  after a SR is closed.}
When an SR is updated, the 311 software automatically populates the 
\texttt{resolution\_action\_update\_date}. Some of these updates 
occur long after the SR is closed. As shown in Figure~\ref{fig:resolution-violin}, 
there are 7,460 SRs that were updated more than 30 days after the 
\texttt{closed\_date}, but less than 730 days, to exclude infeasible 
dates like 1900-01-01. The chart highlights the distribution of these 
post-closure updates, with a notable concentration of updates within 
the 30-to-90-day range. This pattern raises questions about whether 
such delayed updates are standard operational practice or indicative 
of a potential issue requiring further investigation. While the reason 
for these delayed updates is unclear, they may impact the accuracy of 
response time analyses if not properly accounted for.

\subparagraph{Same Created Date \& Closed Date: Zero Durations}
A more prevalent issue occurs when the \texttt{closed\_date} and 
\texttt{created\_date} are exactly the same, down to the second. This 
creates a \textbf{zero duration}, which is also nonsensical. There 
are 191,141 such SRs, representing 3.1\% of all non-blank data. 
Ninety-nine percent of these zero-duration SRs occur in five agencies:
DHMH, DOT, DOB, DSNY, and DEP.. 
This distribution does not mirror the overall SR distribution by 
agency, suggesting an agency-specific issue.

\subparagraph{Created Date or Closed Date(s) at Midnight or Noon}
Another issue found with the \texttt{created\_date} and 
\texttt{closed\_date} fields is the unusually large number of SRs 
created or closed at exactly midnight (00:00:00) or noon (12:00:00), 
down to the second. Normally, SR creation and closure follow the 
work-day schedule, with most SRs created during daylight hours and 
fewer at night or in the early morning. However, there is a 
significantly higher number of SRs closed exactly at midnight and noon, 
as well as a large number created exactly at noon. Specific observations 
include:
\begin{itemize}
    \item 99,779 SRs were created exactly at noon (12:00:00).
    \item 235,347 SRs were closed exactly at midnight (00:00:00).
    \item 105,505 SRs were closed exactly at noon.
\end{itemize}

A unique way to visualize this anomaly is to examine the busiest day 
during this 2-year period (Friday, 2023-09-29). Then aggregate 
the created\_date(s) by minute (with seconds equal to zero), providing
a minute-by-minute look at SR creation on that one day. Note the spike 
at exactly noon (12:00), well beyond the 3\textsigma line.

\begin{figure}[tbp]
    \centering
    \begin{minipage}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SR_created_by_minute_of_busiest_day.pdf}
        \caption{SRs Created Minute-by-Minute on Busiest Day}
        \label{fig:busiestcreated}
    \end{minipage}%
    \begin{minipage}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{2-year-trend-SR_closed_by_minute_of_busiest_day.pdf}
        \caption{SRs Closed Minute-by-Minute on Busiest Day}
        \label{fig:busiestclosed}
    \end{minipage}
\end{figure}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.7\textwidth]{2-year-trend_SR_created_by_minute_of_busiest_day.pdf}
  \includegraphics[width=0.7\textwidth]{2-year-trend-SR_closed_by_minute_of_busiest_day.pdf}
  \caption{SRs Created and closed minute-by-minute on busiest day
    during two year period \jy{give years}.}
  \label{fig:busiest}
\end{figure}

>>>>

\usepackage{comment}
\usepackage{booktabs, textgreek}
\usepackage{subcaption}

@@

The paper is organized as follows. Section~\ref{sec:data} provides 
an overview of the history of the NYC 311 SR system and presents 
a summary of SR counts over a 2-year period from 2022 to 2023. 
Section~\ref{sec:issues} delves into specific data cleansing 
challenges affecting data quality and curation efficiency, 
including structural problems, adherence to the data dictionary, 
and the presence of missing, blank, or N/A entries. This section
 investigates field compliance with reference or acceptable values, 
highlights logical inconsistencies, and examines concerning patterns 
in the data. We explore the balance between precision and accuracy 
and identify duplicate or redundant fields, along with observations 
on the Data Dictionary. Section~\ref{sec:recommendations} offers 
practical recommendations for mitigating or resolving these issues, 
while Section~\ref{sec:discussion} encapsulates key insights and 
discusses the broader implications of our findings.

\section{NYC 311 SR Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system (the biggest expansion in their history), and a 
record 3.23 million service requests in 2023.

@@

\begin{comment}
Despite its success, the 311 system faces several challenges:
data timeliness, accuracy, and consistency, difficulties in
correlating data over long periods, excluding personally
identifiable information (PII), integration with stand-alone systems
at selected NYC agencies, and managing API usage for numerous
third-party users. These challenges are addressed by various agency
open data managers and OTI.
\end{comment}

@@

Our investigation utilizes an SR dataset covering  2-years (2022-2023).
This dataset is approximately 3.7 GB in size and contains over 
6.4 million rows with each row representing one SR. It contains 
41 columns of data. It is used to conduct detailed analysis 
in areas such as data consistency, data validity, redundant fields, 
etc.. Instructions on downloading this file from the 
NYC Open Data Portal are available in the supplemental material.

During the 10-year period from 2014 to 2023,  311 experienced
a 50\% in Service Requests, , reflecting a growing public reliance 
on the system. (As a note, during this 10-year timeframe, the 
population of NYC grew just 0.6\%.). The most significant increase
in activity occurred in 2020, when the number of requests 
surged to 3.23 million during the COVID-19 pandemic, 
illustrating the essential role of the 311 system in supporting 
NYC residents during times of crisis. The steady 
rise in SRs can be at least partially attributed to the increased 
accessibility of the 311 system via online and mobile 
platforms, as well as heightened public awareness of the service. 
While the spike in 2020 was exceptional, the broader trend 
indicates sustained growth in SR volume in future years,  
highlighting the system's evolving role in managing 
both routine city operations and extraordinary events.

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
agency, showing the cumulative percentage of SRs handled by each
agency over a the 2-year timeframe. Note that the
distribution of SRs is heavily concentrated among a few key 
agencies. The six largest agencies are:

\begin{itemize}
    \item New York Police Department (NYPD) - 43\%
    \item Housing Preservation and Development (HPD) - 21\%
    \item New York City Department of Sanitation (DSNY) - 10\%
    \item Department of Transportation (DOT) - 7\%
    \item Department of Environmental Protection (DEP) - 5\%
    \item Department of Parks and Recreation (DPR) - 4\%
\end{itemize}

Collectively these six City agnecies handle over 90\% of the total SR 
volume, indicative of the critical role these agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The NYPD alone accounts for a
substantial portion of the total SRs, underscoring its leading role in
addressing complaints related to public safety. The remaining
10\% of SRs is distributed across 10 other
agencies. This concentration of SRs to the `big six'' agencies 
underscores the necessity for optimal data handling and curation processes 
within these high-volume agencies in order to ensure 
efficiency and responsiveness.

@@

Figure~\ref{fig:SR_complaints}
illustrates how the top 20 complaint types, accounting for 70\% 
of all SRs. Note that the distribution of complaints is skewed, 
with a small number of issues dominating total SR volume. 
Noise-related complaints, including residential, commercial, and 
street noise, when combined make up 22\% of all 
complaints; the most frequent issue in the NYC 311 system. 
Other prominent categories include illegal parking, heat/hot water 
complaints, and blocked driveways. The cumulative percentage curve 
reveals that after the top 20 complaint types, the remaining 
complaint types are spread thinly across the remaining 30\% of SRs,
suggesting that improving responses to the most frequent 
complaints could have an outsized impact on overall service 
efficiency and resident satisfaction. This chart also provides 
insights into the operational pressures faced by City agencies 
responsible for handling these high-volume complaints.

@@

Data cleansing refers to the process of identifying and rectifying
errors, inconsistencies, and inaccuracies within datasets to ensure
they are of high quality and reliable for analysis
\citep{maletic2005data, hosseinzadeh2023data}. The process
typically involves removing duplicate records, handling missing or
incomplete data, correcting mislabeled or inaccurate entries, and
standardizing data formats \citep[e.g.,][]{cody2017cody,
  van2018statistical}. In the context of open data, cleansing is
especially important as open datasets often come from diverse,
uncoordinated sources, leading to variations in data quality,
completeness, and consistency. Without thorough data cleansing, the utility
of open data can be limited and perhaps untrustworhty, affecting 
its reliability for research, policy-making, and innovation. The 
main purpose of cleansing open data is to ensure that it is 
accurate, consistent, and usable across multiple platforms and by 
various stakeholders with a myriad of purpose. Data cleansing improves
the trustworthiness of the data and enables more accurate analysis, 
better decision-making,and the integration of data into machine
learning models or other systems.

Many quality criteria are employed to ensure high-quality data 
processing. One of the primary efforts is data validation, which spans 
several critical checks. For instance, mandatory fields must not be 
left empty, ensuring that key information is always captured. 
Additionally, certain fields must conform to specific data types, 
such as numeric, character, or date formats; typically 
outlined in a Data Dictionary. An essential aspect of validation 
is domain compliance, where fields must adhere to a predefined set of 
values, such as statuses, state names, or zip codes. Structural 
errors also play a significant role in data quality, particularly when 
naming conventions or data entries are inconsistent. Common issues 
include fields that do not appear in the Data Dictionary or the 
inconsistent use of blanks, spaces, NA, N/A, or ``<NA>'' to indicate 
missing data. Furthermore, redundant or irrelevant fields can clutter 
datasets, reducing efficiency. Logical inconsistencies are another 
important consideration, such as related fields that violate expected 
relationships, such as a ``due date'' that precedes the ``created date.'' 
Lastly, the balance between accuracy and precision is crucial, as both 
must be carefully managed to ensure reliable and meaningful data.

Here we identify the presence of issues in the NYC 311 SR open data;
not attempting to solve them but rather to highlight their scope and
magnitued. A solution effort should be undertaken only after further 
investigation as to the why and how such issues came about, 
and a discussion as to whether or not it is an actual \textit{error} or 
holds some other status. Coordination with the originating Agency
is paramount.

@@

	\item There are four date fields (created, closed, action\_updated, due).

@@

	\item In addition to the incident\_address, there are five additional location fields: 
	latitude/longitude, street\_name, landmark, USGS State Plane 
	Coordinate System, and Block \#.

	\item A free-form text field, resolution\_description, which 
	supports 934 characters of input, including commas and special characters
\end{itemize}

The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} 
identifies 41 data columns (fields) and provides related information 
for each. However, there are six additional fields that can be
optionally downloaded. These fields are 
visible in the portal's ``Column manager'' widget, labeled with the prefix 
``@computed\_region.'' The additional fields include:

\begin{itemize}
    \item \texttt{zip\_codes}
    \item \texttt{community\_districts}
    \item \texttt{borough\_boundaries}
    \item \texttt{city\_council\_districts}
    \item \texttt{police\_precincts}
    \item \texttt{police\_precinct}
\end{itemize}

While one can likely infer the meaning and 
derivation of these computed fields, uncertainty remains as to
their derivation and usage. These six fields are not included in
the Data Dictionary, which we have asked to be updated. Our 
analysis shows that these six \texttt{computed} fields have 
significant data validity issues, leading us to treat these six fields 
as experimental and not for suitable for normal use. Accordingly, 
we will not address these fields in our analysis.

\subsection{Missing Data}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to 
determine if SRs were closed before or after their
\texttt{due\_date}, you would be challenged as 99.6\% of the
\texttt{due\_date} field is blank. When counting fields for 
blank or N/A values, they appear to divide into three groups:

\begin{itemize}
    \item \textbf{Mostly Empty} -- 93-99.9\% blank 
    \item \textbf{Partially Empty} -- 40-4\% blank
    \item \textbf{Few/None Empty} -- 2-0\% blank
\end{itemize}

The Mostly Empty category includes such fields as
\texttt{taxi\_company due\_date},
\texttt{pickup\_location}, and \texttt{landmark}.
The Partially Empty includes such fields as
\texttt{location\_type}, \texttt{borough}, 
and \texttt{cross\_street}. And the Few/None Empty includes
\texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. In some analysis efforts, it may be
prudent to inquire as to why some fields are mostly or almost always blank.
Figure~\ref{fig:blank_fields} presents a graphic 
depiction of total empty (blank \& N/As) for each fields illustrating
the grouping into the Most, Partial, and Few categories.

@@

\subsection{Validating Data for Acceptable Values}
\label{sec:domain}
Any analytic effort must ensure that fields containing invalid values 
are identified and isolated from the analysis. Below are the results of 
selected field validations. The latitude and longitude fields were 
found to fall within the geographic boundaries of New York City, and 
the \texttt{unique\_key} field was indeed unique, as required. 
Unfortunately, the Data Dictionary specifies very few domains of 
acceptable values. Nonetheless, the following fields were tested and found 
to comply with their expected domains, as determined by public usage 
and historical datasets:

\begin{itemize}
    \item \texttt{address\_type}
    \item \texttt{status}
    \item \texttt{borough}
    \item \texttt{borough\_boundaries}
    \item \texttt{park\_borough}
    \item \texttt{data\_channel}
    \item \texttt{vehicle\_type}
    \item \texttt{city\_council\_district}
\end{itemize}

But some fields proved problematic as to their 
compliance with a domain of allowable values.

\paragraph{Zip Codes}
\label{sec:zipcodesissues}
All zip codes ( \texttt{incident\_zip}) 
should validate against the USPS database, which contains 
44,173 valid zip codes. The \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,374 errors; 
not insignificant. If you group the invalid entries by 
percentage by Agency, it mirrors the \texttt{overall} 
breakdown of SRs by Agency, potentially 
indicating a systemic problem.

\paragraph{Created and Closed Date}
\label{sec:negativeduration}
For the \texttt{created\_date} and \texttt{closed\_date} fields, one 
might expect that these fields are automatically populated by the SR 
application software when setting an SR status to ``new'' or ``closed.'' 
Unfortunately, this does not seem to be the case, as several anomalies 
exist in the created and closed date fields. Here are some of the errors:

\begin{itemize}
    \item SRs with a \texttt{closed\_date} that occurs before the 
    \texttt{created\_date} indicating a negative duration.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} in 
    the far distant past, e.g. 1900.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} that 
    match to the second indicating a zero duration.
    \item A large number of SRs closed and/or created exactly at midnight 
    or noon, to the second.
\end{itemize}

\subparagraph{Closed before Created: Negative Duration}
Citizens, NYC Government Officials, and Agencies use the created and 
closed dates to measure the ``duration'' of SRs, which reflects Agency 
responsiveness. While duration is not directly present in the dataset, 
it can be easily computed as the difference between
\texttt{(closed\_date} and \texttt{created\_date}.  Duration is one of 
most used fields that is constantly analyzed, such as determining if 
one NYC borough receives faster response for ceratin 
complaints. There are 12,450 SRs where the \texttt{closed\_date} precedes the 
\texttt{created\_date}, generating nonsensical ``negative durations.'' 
Though this represents only 0.2\% of the dataset, these errors can 
significantly impact response time analyses. Eight SRs with extremely 
large negative durations ($-$730 days), all originating from the 
Department of Homeless Services (DHS); all contain an entry 
of ``1900-01-01'' as the \texttt{closed\_date}. This results 
in negative durations exceeding $-$44,601 days (or 122 years). Many
other SRs have negative durations in excess of 300+ days. Such 
anomalies, though rare, can significantly skew statistical results if 
not addressed during data cleansing. As a result, these SR rows are 
removed from the analysis. Here is a sample:

@@

Excluding the extreme negative values, 
Figure~\ref{fig:negative-duration-violin} shows the broad spread of 
negative-duration SRs. While there are few outliers, the magnitude 
of the negative durations is troubling and can produce bizarre
and unwanted analytical results.

@@

Table~\ref{tab:largest-errors} summarizes the 
negative durations of the largest magnitude (in days). The data 
suggests that the negative duration issue is predominantly a problem 
within the Department of Transportation (DOT), where 95\% of these 
errors occur.

@@

A similar issue occurs when resolution action update date is after the
closing date.  When an SR is updated, the 311 software 
automatically populates the \texttt{resolution\_action\_update\_date}. 
While it is certainly possible to update an SR after it is closed, 
some of these updates seem to be well beyond an expected 
range, some over 2 years later. As shown in 
Figure~\ref{fig:resolution-violin}, there are 10,265 SRs that 
were updated more than 30 days after the 
\texttt{closed\_date}, but less than 730 days, cutoff used to 
exclude infeasible dates such as 1900-01-01. The chart 
highlights the distribution of these post-closure updates, with 
a notable concentration of updates within 
the 30-to-90-day range. This pattern raises questions about whether 
such delayed updates are standard operational practice or indicative 
of a potential issue requiring further investigation.

@@

\subparagraph{Identical Created Date \& Closed Date: Zero Durations}
A more prevalent issue occurs when the \texttt{closed\_date} and 
\texttt{created\_date} are exactly the same, down to the second. This 
creates a \textbf{zero duration}, which is nonsensical. There 
are 163,720 such SRs, representing 2.6\% of all non-blank data. 
Ninety-nine percent of these zero-duration SRs occur in five agencies:
DHMH, DOT, DOB, DSNY, and DEP. This distribution does not 
mirror the overall SR distribution by Agency, suggesting 
an agency-specific issue.

\subparagraph{Created Date or Closed Date(s) at Midnight or Noon}
Another issue found with the \texttt{created\_date} and 
\texttt{closed\_date} fields is the unusually large number of SRs 
created or closed at exactly midnight (00:00:00) or noon (12:00:00), 
down to the second. Normally, SR creation and closure follows the 
work-day schedule, with most SRs created during daylight hours and 
fewer at night or in the early morning. However, there is a 
significantly higher number of SRs closed exactly at midnight, 
as well as a large number created exactly at midnight.

\begin{figure}[tbp]
    \centering
    \begin{subfigure}{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_created_on_the_hour.pdf}
        \caption{SRs Created by Hour of the Day}
        \label{fig:busiestcreated}
    \end{subfigure}
    \par\medskip
    \begin{subfigure}{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_closed_on_the_hour.pdf}
        \caption{SRs Closed by Hour of the Day}
        \label{fig:busiestclosed}
    \end{subfigure}
    \caption{Comparison of SRs Created and Closed by Hour of the Day}
    \label{fig:stacked}
\end{figure}
